---
title: "New York City Dog Bite Analysis"
author: "Ian Young"
output: 
  pdf_document:
    toc: yes
    fig_caption: yes
---

\newpage

# Load Packages
```{r, warning=FALSE, message=FALSE}
library(tidyverse); library(showtext); library(scales)
font_add_google("Roboto Condensed")
showtext_auto()
library(httr2)
library(gt)
```

# Connect to API & Compile Dataset

The httr2 and purrr packages are used to connect to an API from New York City and 
convert dog bite data for the years 2015 to 2022 from an unstructured format into 
a data frame.
```{r}
# Collect data on dog bites in NYC
dog_bites <- 
  request("https://data.cityofnewyork.us/resource/rsgh-akpg.json") |> 
  req_url_path_append("?$limit=30000") |> # change call limit to collect all records
  req_perform() |> 
  resp_body_json() |> 
  map_dfr(~ as_tibble(.)) # convert lists into one tibble
```

## Initial Data Cleaning

An initial organizing and cleaning of the data takes place using . This involves converting 
the date column that is stored as character data type to a date type, and adding a 
individual column for year, month, and day. This is done using the lubridate package. 
The uniqueid column was not recorded properly in the initial data set, and a more helpful 
uniqueid column is created. Lastly, the species column is dropped as it provides no 
information (all values in it are "DOG") and the columns are rearranged.
```{r}
# Correct unique ID, adjust/create date columns, select relevant variables
dog_bites_clean <- 
  dog_bites |> 
  arrange(dateofbite) |> 
  mutate(uniqueid = row_number(),
         dateofbite = as_date(dateofbite),
         biteyear = year(dateofbite),
         bitemonth = month(dateofbite),
         biteday = day(dateofbite)) |> 
  select(uniqueid, dateofbite, biteyear, 
         bitemonth, biteday, zipcode, 
         borough, breed, gender,
         age, spayneuter)
```

\newpage

# Background & Initial Exploration
Below is a sample of 10 observations from the initial cleaned dog bite data. 
There are a total of 26,127 observations and 11 variables (3 data columns not displayed 
in the table below). Each observation represents a single unique dog bite incident 
collected by NYC between the start of 2015 and the end of 2022. The data was collected 
online, by mail/fax, and by phone by the health departments animal bite unit. As 
a result columns like zipcode, breed, and age all contain a range of values including 
missing data. For the spayneuter column FALSE represents dogs that have not had their 
reproductive organs surgically removed, as well as cases where it is unknown if the 
dog involved in the incident has or has not been neutered. Note: The skimr package 
and function skim are not rendered below, they were used in gathering a quick general 
understanding of the data.
```{r}
# Sample of 10 observations
set.seed(30)
dog_bites_clean |> 
  select(-biteyear, -bitemonth, -biteday) |> 
  sample_n(10) |> 
  gt() |> 
  tab_header(
    title = md("**NYC Dog Bite Reports 2015-2022**"),
    subtitle = md("Sample of 10 Observations")
  ) |> 
  tab_source_note(
    source_note = md("Source: 
    https://data.cityofnewyork.us/Health/DOHMH-Dog-Bite-Data/rsgh-akpg/about_data")
  ) |> 
  tab_footnote(
    footnote = md("**FALSE** represents dogs that have not been neutered **OR** 
                  it was unknown."),
    locations = cells_column_labels(
      columns = spayneuter
    )
  )

# Method for quick glimpse of the data
# skimr::skim(dog_bites_clean)
```

## Further Data Cleaning

# Analysis

## Mapping

## Group comparisons

## Forecasting

## Predicting

# Reference
```{r}
# clean up age column, zip code column, pull in long/latitude?

# format gt table
# group data by month and pull potential predictors of dog bites per month

# ts <- 
#   test |> 
#   group_by(biteyear, bitemonth) |> 
#   count() |> 
#   mutate(date = make_date(biteyear, bitemonth)) |> 
#   ungroup() |> 
#   select(date, n)
# 
# ts |> 
# ggplot(aes(x = date, y = n)) +
#   geom_line() + 
#   scale_x_date(date_labels = "%Y", date_breaks = "1 year") +
#   labs(x = "Date", y = "Dog Bites", title = "NYC Dog Bites Over Time") +
#   theme_minimal()

# nyc_weather <-
#   request("https://api.weather.gov") |> 
#   req_url_path_append("points", 
#                       "40.74,-73.98"
#                       ) |> 
#   req_perform() |> 
#   resp_body_json() |> 
#   pluck("properties", "forecast") |> 
#   glimpse()
#   map_dfr(~ as_tibble(.))
```
  
  

