---
title: "New York City Dog Bite Analysis"
author: "Ian Young"
output: 
  pdf_document:
    toc: yes
    fig_caption: yes
---

\newpage

# Load Packages
```{r, warning=FALSE, message=FALSE}
library(tidyverse); library(showtext); library(scales)
font_add_google("Roboto Condensed")
showtext_auto()
library(httr2)
library(gt)
```

# Connect to API & Compile Dataset

The httr2 and purrr packages are used to connect to an API from New York City and 
convert dog bite data for the years 2015 to 2022 from an unstructured format into 
a data frame.
```{r}
# Collect data on dog bites in NYC
dog_bites <- 
  request("https://data.cityofnewyork.us/resource/rsgh-akpg.json") |> 
  req_url_path_append("?$limit=30000") |> # change call limit to collect all records
  req_perform() |> 
  resp_body_json() |> 
  map_dfr(~ as_tibble(.)) # convert lists into one tibble
```

## Initial Data Cleaning

An initial organizing and cleaning of the data takes place. This involves converting 
the date column that is stored as character data type to a date type, and adding a 
individual column for year, month, and day. This is done using the lubridate package. 
The uniqueid column was not recorded properly in the initial data set, and a more helpful 
uniqueid column is created. Lastly, the species column is dropped as it provides no 
information (all values in it are "DOG") and the columns are rearranged.
```{r}
# Correct unique ID, adjust/create date columns, select relevant variables
dog_bites_clean <- 
  dog_bites |> 
  arrange(dateofbite) |> 
  mutate(uniqueid = row_number(),
         dateofbite = as_date(dateofbite),
         biteyear = year(dateofbite),
         bitemonth = month(dateofbite),
         biteday = day(dateofbite)) |> 
  select(uniqueid, dateofbite, biteyear, 
         bitemonth, biteday, zipcode, 
         borough, breed, gender,
         age, spayneuter)
```

\newpage

# Background & Initial Exploration
Below is a sample of 10 observations from the initial cleaned dog bite data. 
There are a total of 26,127 observations and 11 variables (3 data columns not displayed 
in the table below). Each observation represents a single unique dog bite incident 
collected by NYC between the start of 2015 and the end of 2022. The data was collected 
online, by mail/fax, and by phone by the health departments animal bite unit. As 
a result columns like zipcode, breed, and age all contain a range of values including 
missing data. For the spayneuter column FALSE represents dogs that have not had their 
reproductive organs surgically removed, as well as cases where it is unknown if the 
dog involved in the incident has or has not been neutered. Note: The skimr package 
and function skim are not rendered below, they were used in gathering a quick general 
understanding of the data.
```{r}
# Sample of 10 observations
set.seed(30)
dog_bites_clean |> 
  select(-biteyear, -bitemonth, -biteday) |> 
  mutate(dateofbite = format(dateofbite, "%Y-%m-%d")) |> 
  sample_n(10) |> 
  gt() |> 
  tab_header(
    title = md("**NYC Dog Bite Reports 2015-2022**"),
    subtitle = md("Sample of 10 Observations")
  ) |> 
  tab_source_note(
    source_note = md("Source: 
    https://data.cityofnewyork.us/Health/DOHMH-Dog-Bite-Data/rsgh-akpg/about_data")
  ) |> 
  tab_footnote(
    footnote = md("**FALSE** represents dogs that have not been neutered **OR** 
                  it was unknown."),
    locations = cells_column_labels(
      columns = spayneuter
    )
  )
```

## Further Cleaning 

After the inital data cleaning a few more steps occur to prepare the data for analysis. 
Using the zipcodeR package data on census and geographical data are joined with 
the dog bite data. This data reflects 2010 census and zip code data. 

The data contains values input from outside of New York City and has missing location 
data values. For simplicity sake records outside of New York City and the records 
missing location data are filtered out.

The breed column contains open text responses with a wide variety of inputs. Some 
simple steps are taken to format including making text lower case, removing white 
spaces, and removing trailing non letter characters.

Lastly, the age column contains a variety of open text responses reporting age of 
the dog involved in the incident. These inputs are taken and transformed to represent 
how old in months the dog was at the time of the incident.

```{r, warning=FALSE, message=FALSE}
# Load in U.S. ZIP code data from zipcodeR to join with dog bite data
library(zipcodeR)

zip_db <- 
  zip_code_db |> 
  select(zipcode, lat, lng, major_city, state, 
         county, radius_in_miles, population, population_density, 
         land_area_in_sqmi, water_area_in_sqmi, housing_units, 
         occupied_housing_units, median_home_value, median_household_income)

# Join dog bite and zip code data
dog_bites_full <- 
  dog_bites_clean |> 
  left_join(zip_db, by = "zipcode")

# Rearrange columns
dog_bites_full <- 
  dog_bites_full |> 
  relocate(zipcode, .after = spayneuter) |> 
  relocate(borough, .after = lng) |> 
  relocate(land_area_in_sqmi, water_area_in_sqmi, .after = radius_in_miles)

# Filter missing location data and down to NYC records
nyc_counties <- c("Bronx County", "Kings County", "New York County", 
                  "Queens County", "Richmond County")

dog_bites_full <- 
  dog_bites_full |> 
  drop_na(lat) |> 
  filter(state == "NY") |> 
  filter(county %in% nyc_counties)

# Clean up breed column text
dog_bites_full <- dog_bites_full %>%
  mutate(breed = str_to_lower(breed), # change text to lower case
         breed = str_trim(breed), # remove start and end white space
         breed = str_squish(breed), # internal white space replaced with single space
         breed = str_remove(breed, "[^a-z]+$") # remove non letter trailing characters
         )

# Clean up age column

```

# Analysis

## Exploratory Data Analysis
```{r}
# Descriptive Statistics/Plots
```

## Mapping Dog Bites In NYC
```{r, warning=FALSE, message=FALSE}
# Load mapping package
library(sf)

# Read in shape file
shape_nyc <- st_read("Borough Boundaries.geojson", quiet = TRUE)

# Begin creation of map
ggplot() +
  geom_sf(data = shape_nyc, aes(fill = boro_name), show.legend = FALSE) +
  theme_void()
```

## Forecasting Dog Bites In NYC
```{r}
# Create data frame for time series analysis
dog_ts <-
  dog_bites_clean |>
  group_by(biteyear, bitemonth) |>
  count() |>
  mutate(
    season = case_when(
    bitemonth %in% c(3, 4, 5) ~ "spring",
    bitemonth %in% c(6, 7, 8) ~ "summer",
    bitemonth %in% c(9, 10, 11) ~ "fall",
    bitemonth %in% c(12, 1, 2) ~ "winter"), 
    date = make_date(biteyear, bitemonth)
    ) |>
  ungroup() |>
  select(date, season, n)
  
# Initial plot of time series data
dog_ts |>
mutate(season = fct_relevel(season, "spring", "summer", "fall", "winter")) |> 
ggplot(aes(x = date, y = n)) +
  geom_line() +
  geom_point(aes(color = season), size = 3, alpha = 0.7, shape = 20) +
  scale_x_date(date_labels = "%Y", date_breaks = "1 year") +
  labs(x = "Date", 
       y = "Dog Bites", 
       title = "NYC Dog Bites 2015-2022",
       subtitle = "Dog bites regularly spike in the summer months with 2020 outlier.",
       caption = "Source: NYCOpenData") +
  scale_color_manual(values = c("spring" = "darkgreen", 
                                "summer" = "firebrick1", 
                                "fall" = "goldenrod4", 
                                "winter" = "dodgerblue4")) +
  theme_bw() +
  theme(
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 10),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 10),
    plot.caption = element_text(size = 8),
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.title = element_blank(),
    legend.text = element_text(size = 10)
    )
```
